---
title: "mpathr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mpathr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mpathr)
```

The main goal of `mpathr` is to provide functions to import data from the m-Path
platform, as well as provide functions for common manipulations for
ESM data.

## Importing m-Path data

To show how to import data using `mpathr`, we provide example data within 
the package:

```{r show m-Path example data}
mpath_example()
```
As shown above, the package comes with an example of the `basic.csv` that can be
exported from the m-Path platform. \n

To read this data into R, we can use the `read_mpath` function. We will also 
need a path to the meta data. 

The main advantage of using `read_mpath`, as opposed to other functions like 
`read.csv`, is that `read_mpath` uses the meta data to correctly interpret the
data types. Furthermore it will also automatically convert columns that store 
multiple responses into lists. For a response with multiple options like 1,4,6,
`read_mpath` will store a list with each number, which facilitates further 
preprocess these responses.\n

We can obtain the paths to the example basic data and meta data 
using the `mpath_example` function: 

```{r use read_mpath}
# find paths to example basic and meta data:
basic_path <- mpath_example(file ='example_basic.csv')
meta_path <- mpath_example('example_meta.csv')

# read the data
data <- read_mpath(file = basic_path,
                   meta_data = meta_path)
```

### Note: saving m-Path data
As explained above, the resulting data frame will contain columns with lists,
which can be problematic when saving the data. To save the data, we suggest
using either `fwrite` from the `data.table` package, or the `save` function. 

If you want to save the data as a 'csv' to use it in another program, use
`fwrite` from the `data.table` package:

```{r write data as csv}
library(data.table)

fwrite(data, 
       file = 'data.csv', # specify path
       sep = ';', # separator between columns
       sep2 = c('', ',', '')) # separator within list columns
```

Otherwise, if the data will be used in R, we suggest saving it as an R object:
```{r write data as an R object}
# save the data as an R object
save(data, 
     file = 'data.RData') # specify path
```

## Obtaining response rates

### response_rate function

Some common operations that are done on ESM data have to do with the 
participants' response rate. We provide a function `response_rate` that 
calculates the response_rate per participant for the entire duration of the 
study, or for a specific time frame.\n

This function takes as argument a `valid_col`, that takes a logical column that 
stores whether the beep was answered by the participant, or not, as well as a 
`participant_col`, that identifies each separate participant.

We will show how to use this function with the `example_data`, that contains data from the same 
study as the `example_basic.csv` file, but after some cleaning.\n

```{r calculate response rate}
data(example_data)

response_rates <- response_rate(data = example_data,
                               valid_col = answered,
                               participant_col = participant)
```

The function returns a data frame with:\n

* The `participant` column, as specified in `participant_col`
* The `number_of_beeps` used to calculate the response rate.
* The `response_rate` column, which is the proportion of valid responses 
  (specified in `valid_col`) per participant.

The output of this function can further be used to identify participants with 
low response rates:

```{r show low response rates}
response_rates[response_rates$response_rate < 0.5,]
```

We could also be interested in seeing the participants' response rate during
a specific period of time (for example, if we think a participant's compliance
significantly dropped a certain date). In this case, we should supply the 
function with the (otherwise optional) argument `time_col`, that should contain
times stored as 'POSIXct' objects, and specify the date period that we are
interested in (in the format 'yyyy-mm-dd' or 'yyyy/mm/dd'):

```{r calculate response rate after 15th of May 2024}
response_rates_after_15 <- response_rate(data = example_data,
                               valid_col = answered,
                               participant_col = participant,
                               time_col = sent,
                               period_start = '2024-05-15')
```

This will return the participant's response rate after the 15th of May 2024. 

```{r show low response rates after 15th of May 2024}
response_rates_after_15
```

### plot_response_rate function

We also suggest a way to plot the participant response rates, to identify 
patterns like response rates dropping over time. For this, we provide the `plot_response_rate` function.

```{r plot response rate, fig.width=7, fig.height=5}
plot_response_rate(data = example_data,
            time_col = sent,
            participant_col = participant,
            valid_col = answered)
```
Note that the resulting plot can be further customized using the `ggplot2`
package.

```{r customize plot response rate plot, fig.width=7, fig.height=5}
library(ggplot2)

plot_response_rate(data = example_data,
            time_col = sent,
            participant_col = participant,
            valid_col = answered) +
  theme_minimal() +
  ggtitle('Response rate over time') +
  xlab('Day in study')
```

